Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-12 Manuscript/manuscript.tex
Added a few statements to stoichiometry section, now it shall be ready for re-writing it into a coherent text.

This section is rough and will likely require editing. 
Data -- up to noted exceptions -- shall be all there, however. 
\todo{SAMULI: I think the we should make clear that the current binding
  constants in, e.g., Marsh's handbook are based on the headgroup order parameter
  changes intepreted with ternary complex binding model. I.e. the experimental raw data
  is excatly the same as we have in this work. Our simulations enable much more versatile interpretation
  of the binding phenomena. I think that this is one the main points of this work, so this comment
  applies to the previous section, introduction, abstract and conclusions as well.
}

@@

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
The dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Such a finding is based on the calculated
number of contacts per lipid between \ce{Ca^{2+}} and POPC oxygen atoms, 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
in which a contact between the two groups is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ (robustness of this definition discussed in reference~\citenum{javanainen17}). 
the average number of contacts between \ce{ca^{2+}} and any oxygen atom in popc is 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.47 $,    % = 59.8/128 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.45 $,    % = 57.2/128 . 
and if only carbony oxygen atoms are considered,
this value is merely
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.02 $.    % = (59.8-57.2)/128 . 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
which are easier to interpret.

Internal side-note: 
However, $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$ is increased 
when carbonyl dipoles are scaled only with 0.945 
(roughly done through charge redistribution)
 to : $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} \approx 0.08 $.    % = (70.-59.8)/128 . 
This does not change the dominance of phosphates, which still interact more than 6x times more. That's good.
In addition, the probability of states, containing sn-1:O increases from 6\% to 10\%.
The residence times change to 75ns (90\% of bound time) and max is 185ns.
That's not much, too. It certainly is not a game changer, I'd even call it to be within accuracy. 
The slight problem may be that with this tweak, there's approx. 20\%  more cations bound to the membrane,
which means, I'd need to tune the sigma parameter a bit. 
Not $f_q$ -- that's set by the simulations with the surfactant. 
That could detune the membrane structure \dots
I think that's within the accuracy.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From a recent theoretical work with long simulations 
(Fig. S17 in \citenum{javanainen17} -- C36 with MK Ca2+, any other model, either different cation or Slipids/C36, just such calcium)
this time can be roughly estimated -- get the simulation from \url{https://doi.org/10.5281/zenodo.259376} and analyze it -- 
in the order of $1$--$10\,\mu\mathrm{s}$~\cite{javanainen17}. 
This is in contrast to our model, ECC-lipids, 
which estimates the residence time of calcium bound to a POPC membrane to be 
smaller than $53\,\mathrm{ns}$ for 90\% of the time a cation is in a membrane-bound state. 
i.e. at least two orders of magnitude lower than previous estimates.
In addition, the longest observed residence time in our simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.
This is unfortunately not the case of many theoretical works,
where the residence time cannot be estimated,
or the simulation has not even decorrelated from it's 
initial condition with a population of bound/unbound cations \cite{catte16}, 
because it is limited by the simulation length, 
even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
Increasing time scales to the order of $\mu\mathrm{s}$ 
doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent. 
A histogram of residence times from our simulations is in Fig.~S\ref{fig:hist_residence_times} in SI. 
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
Add data for Na+. }

>>>>

This section is rough and will likely require editing. 
Data -- up to noted exceptions -- shall be all there, however. 
\todo{SAMULI: I think the we should make clear that the current binding
  constants in, e.g., Marsh's handbook are based on the headgroup order parameter
  changes intepreted with ternary complex binding model. I.e. the experimental raw data
  is excatly the same as we have in this work. Our simulations enable much more versatile interpretation
  of the binding phenomena. I think that this is one the main points of this work, so this comment
  applies to the previous section, introduction, abstract and conclusions as well.
  JOE: agree, this would be a good statement in general. I'll try to put it into the text.
}

@@

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
The dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Such a finding is based on the calculated
number of contacts per lipid between \ce{Ca^{2+}} and POPC oxygen atoms, 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
in which a contact between the two atoms is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ 
(robustness of this definition discussed in reference~\citenum{javanainen17}). 
The average number of contacts between \ce{Ca^{2+}} and any oxygen atom in POPC is 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.47 $,    % = 59.8/128 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.45 $,    % = 57.2/128 . 
and if only carbonyl oxygen atoms are considered,
this value is merely
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.02 $.    % = (59.8-57.2)/128 . 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
which are easier to interpret.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From a recent theoretical work with long simulations 
(Fig. S17 in \citenum{javanainen17} -- C36 with MK Ca2+, any other model, either different cation or Slipids/C36, just such calcium)
this time can be roughly estimated %-- get the simulation from \url{https://doi.org/10.5281/zenodo.259376} and analyze it -- 
in the order of $1\,\mu\mathrm{s}$~\cite{javanainen17}.

Improve the above statement so that it contains the reason why I estimate it to this value.
Quick result (Matti C36 with CaClMK, 800ns total):
90\% of bound residence time is smaller than 787 ns, 
The longest is 800 ns (the simulation length)
and it is only less than
60\% of bound residence time is smaller than 400 ns -- half of the simulation length 
and it is only less than
50\% of bound residence time is smaller than 290 ns.
Apparently the exchange of calcium between membrane and solvent 
is apparently limited by the length of the simulation.

This is in contrast to our model, ECC-lipids, 
which estimates the residence time of calcium bound to a POPC membrane to be 
smaller than $53\,\mathrm{ns}$ for 90\% of the time a cation is in a membrane-bound state. 
i.e. at least two orders of magnitude lower than previous estimates.
In addition, the longest observed residence time in our simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.
This is unfortunately not the case of many theoretical works,
where the residence time cannot be estimated,
or the simulation has not even decorrelated from it's 
initial condition with a population of bound/unbound cations \cite{catte16}, 
because it is limited by the simulation length, 
even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
Increasing time scales to the order of $\mu\mathrm{s}$ 
doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.

Interestingly, \ce{Na+} cations exhibit much more rapid exchange between membrane and solvent than calcium.
The residence time of sodium bound to a POPC membrane is
smaller than $2\,\mathrm{ns}$ for 90\% of the time it is bound to the membrane 
with the longest measured bound time being $13\,\mathrm{ns}$. 
This is an order of magnitude faster exchange than for calcium cations.

A histogram of residence times from our simulations is in Fig.~S\ref{fig:hist_residence_times} in SI. 
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
 }

Internal side-note: 
However, $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$ is increased 
when carbonyl dipoles are scaled only with 0.945 
(roughly done through charge redistribution)
 to : $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} \approx 0.08 $.    % = (70.-59.8)/128 . 
This does not change the dominance of phosphates, which still interact more than 6x times more. That's good.
In addition, the states that have a contribution from carbonyl oxygens also contain a bound phosphate, 
so it can be thought of rather as a stabilizing moiety. 
In addition, the probability of states, containing sn-1:O increases from 6\% to 10\%.
The residence times change to 75ns (90\% of bound time) and max is 185ns.
That's not much, too. It certainly is not a game changer, I'd even call it to be within accuracy. 
The slight problem may be that with this tweak, there's approx. 20\%  more cations bound to the membrane,
which means, I'd need to tune the sigma parameter a bit (increase the scaling factor $f_\sigma$). 
That could detune the membrane structure \dots
Not $f_q$ -- that's set by the simulations with the surfactant. Good.
I think that such a binding detail is within the accuracy of the proposed simple correction.
It would be worth noting that the model may underestimate slightly the contribution from carbonyls due to this. 
But we can't say to what extent\dots
